Simplified integrated nested Laplace approximation for extended latent Gaussian models

### 2022-09-14 src/docs_paper/paper.Rmd
Iterate draft

---
title: Simplifying Integrated nested Laplace approximation with adaptive Gaussian Hermite quadrature
author:
  - name: Adam Howes
    email: ath19@ic.ac.uk
    affiliation: A
    corresponding: ath19@ic.ac.uk
address:
  - code: A
    address: Department of Mathematics, Imperial College London
abstract: |
  |  **Background:**
  |
  |  **Methods:**
  |
  |  **Results:**
  |
  |  **Conclusions:**
author_summary: |
  Author summary
bibliography: citations.bib
output:
  rticles::elsevier_article:
    number_sections: true
    includes:
      in_header: preamble.tex
---

@@

The `TMB` package \citep{kristensen2016tmb} is gaining popularity in spatial statistics as a flexible alternative to `R-INLA` for fitting latent Gaussian models \citep{osgoodzimmerman2021statistical}.

@@

Integrated nested Laplace approximation (INLA) \citep{rue2009approximate} is an approximate Bayesian inference method which uses the Laplace approximation and numerical integration.
INLA is designed for use with latent Gaussian models (LGMs) of the form
\begin{alignat}{2}
&\text{(Observations)}     &        y_i &\sim p(y_i \, | \, x_i, \btheta), \quad i = 1, \ldots, n, \label{eq:data} \\
&\text{(Latent field)}     &        \x &\sim \mathcal{N}(\x \, | \, \mathbf{0}, \mathbf{Q}(\btheta)^{-1}), \label{eq:process} \\
&\text{(Parameters)}       & \qquad \btheta &\sim p(\btheta), \label{eq:parameters}
\end{alignat}
where $\text{dim}(\y) = \text{dim}(\x) = n$ and $\text{dim}(\btheta) = m$, and $m < n$.
The joint posterior of $(\x, \btheta)$ is given by
\begin{equation}
  p(\x, \btheta \, | \, \y) 
  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right). \label{eq:lgmjointposterior}
\end{equation}
Rather than approximating the above full posterior, the INLA method instead approximates the posterior marginals of each latent random variable $x_i$ and parameter $\theta_j$ given by
\begin{align}
  p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\
  p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2}
\end{align}
An approximation is made to each of the two quantities, $p(\btheta \, | \, \y)$ and $p(x_i \, | \, \btheta, \y)$, nested inside the above integrals: (i) $p(\btheta \, | \, \y) \approx \tilde p(\btheta \, | \, \y)$ and (ii) $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$, which we discuss in turn.

@@

The posterior marginal of the parameters $p(\btheta \, | \, \y)$ appears in both Equations \eqref{eq:inla1} and \eqref{eq:inla2}.
This distribution is approximated by $\tilde p(\btheta \, | \, \y)$ and represented by a set of $K$ integration points $\{ \btheta^{(k)} \}$ and area-weights $\{ \Delta^{(k)} \}$.
The first step is to rewrite $p(\btheta \, | \, \y)$ as
\begin{equation}
  p(\btheta \, | \, \y) 
  = \frac{p(\x, \btheta \, | \, \y)}{p(\x \, | \, \btheta, \y)}
  \propto \frac{p(\y, \x, \btheta)}{p(\x \, | \, \btheta, \y)}. \label{eq:hyperpost}
\end{equation}
Approximation (i) then uses a Gaussian approximation
\begin{equation}
p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y) \triangleq \mathcal{N}(\x \, | \, \hat{\bm{\mu}}(\btheta), \hat{\mathbf{Q}}(\btheta)^{-1}) \label{eq:gaussianx}
\end{equation}
to the denominator of Equation \ref{eq:hyperpost}.
This approximation is accurate as the Gaussian prior on the latent field $\x$ makes the posterior distribution
\begin{equation}
p(\x \, | \, \btheta, \y) \propto
\exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right)
\end{equation}
close to being Gaussian since $\y$ is generally not that informative and the observation distribution $p(\y \, | \, \x, \btheta)$ is usually well-behaved \citep{blangiardo2015spatial}.
<!-- If $\y$ is uninformative about $\x$ then how can we hope for useful inference? Perhaps what the author (Marta) meant is that the Gaussian distribution is sufficiently flexible to accommodate the updates made by $\y$. -->

As $p(\btheta \, | \, \y)$ does not depend on $\x$, any value may be chosen to evaluate the right hand side of Equation \ref{eq:hyperpost}.
As such, taking $\x = \hat{\bm{\mu}}(\btheta)$, the value where the Gaussian approximation is most accurate, gives the final approximation as
\begin{equation}
\tilde p(\btheta \, | \, \y) \propto \frac{p(\y \, | \, \x, \btheta) p(\x \, | \, \btheta) p(\btheta)}{p_G(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)} = \frac{p(\y, \hat{\bm{\mu}}(\btheta), \btheta)}{\det(\hat{\bm{Q}}(\btheta))^{1/2}}, \label{eq:hypermarginal}
\end{equation}
where the final equality is because $p_G(\x \, | \, \btheta, \y)$ is evaluated at its mode $\hat{\bm{\mu}}(\btheta)$.

@@

Having utilised used the approximation $p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y)$ in Section~\ref{sec:approxi}, a natural approach, and that taken by \citet{rue2007approximate}, is to marginalise this distribution directly
\begin{equation}
\tilde p(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, \hat{\mu}_i(\btheta), 1 / \hat{q}_i(\btheta)),
\end{equation}
where the marginal mean $\hat{\mu}_i(\btheta)$ and precision $\hat{q}_i(\btheta)$ are recovered directly from $\hat{\bm{\mu}}(\btheta)$ and $\mathbf{Q}^\star(\btheta)$ respectively.
Although this approximation is fast, it tends not to be accurate, as it involves evaluating the Gaussian approximation away from its mode.
As a result, although this method is available in \texttt{R-INLA} it is generally not advised.
Instead, \citet{rue2009approximate} propose two methods, a Laplace approximation and a simplified version which is less computationally demanding.
The full Laplace approximation is
\begin{align}
  p(x_i \, | \, \btheta, \y) &= p(\x \, | \, \btheta, \y) \times \frac{1}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\
  &= \frac{p(\x, \btheta \, | \, \y)}{p(\btheta \, | \, \y)} \times \frac{1}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\
  &\propto \frac{p(\x, \btheta \, | \, \y)}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\
  &\approx \frac{p(\x, \btheta \, | \, \y)}{p_G(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat{\bm{\mu}}_{-i}(x_i, \btheta)} 
  = \tilde p_{LA}(x_i \, | \, \btheta, \y), \label{eq:laplaceapproximation}
\end{align}
where $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ is the Gaussian approximation to $\x_{-i} \, | \, x_i, \btheta, \y$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is its modal configuration.\footnote{This notation is somewhat dangerous as $\hat{\bm{\mu}}(\btheta)$ is the mode of the Gaussian approximation to the full latent field given $\btheta$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is not the same as $\hat{\bm{\mu}}_{-i}(\btheta)$.}
The set of distributions $\{p(\x_{-i} \, | \, x_i, \btheta, \y)\}_{i = 1}^n$ are usually reasonably Gaussian so this approximation tends to work well.
However, the Gaussian approximation $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ must be recomputed for each value of $x_i$ and $\btheta$, which is often computationally prohibitive.
Two modifications to Equation \eqref{eq:laplaceapproximation} are proposed by \citet{rue2009approximate} to reduce the computational cost:
\begin{enumerate}
\item
Avoiding having to find the mode via optimisation by using the approximation $\hat{\bm{\mu}}_{-i}(x_i, \btheta) \approx \mathbb{E}_{p_G(\x \, | \, \btheta, \y)}(\x_{-i} \, | \, x_i)$
\item
As only those $x_j$ close to $x_i$ should have an impact on the marginal of $x_i$, then by selecting some subset $R_i(\btheta)$ of nodes $j$ to impact $j$ the matrix which needs to be factorised can be reduced in dimension to be $| R_i(\btheta) | \times | R_i(\btheta) |$ rather than $n \times n$
\end{enumerate}

@@

# Template Model Builder \label{sec:tmb}

Template Model Builder (TMB) \citep{kristensen2016tmb} is an \textsf{R} package for fitting random effect models, also known as latent variable models, hierarchical models or a host of other names.
In \texttt{TMB} inference is based upon optimisation of a target function.
This makes it very flexible, and able to handle non-linear, non-Gaussian random effect models.

The approach of \texttt{TMB} is inspired by the AD Model Builder (ADMB) package \citep{fournier2012ad}.
The "AD" in ADMB is automatic differentiation, a technique for calculating derivates of functions by repeated application of the chain rule.
AD is popular in machine learning \citep{baydin2017automatic}, for example as the basis for backpropagation algorithm and is beginning to gain popularity in statistics, including as a part of Stan \citep{carpenter2017stan}.
\texttt{TMB} uses the derivatives from AD for multiple purposes including calculation of the Hessian used in Gaussian approximations and for numerical optimisation routines.

## Statistical framework

Consider unobserved latent random effects $\x \in \mathbb{R}^n$ and parameters $\btheta \in \mathbb{R}^m$.\footnote{\citet{kristensen2016tmb} use the notation $u$ for random effects and $\theta$ for parameters. We aim for consistency with Section~\ref{sec:inla}.}
Let $\ell(\x, \btheta) \triangleq - \log p(\y \, | \, \x, \btheta)$ be the negative joint log-likelihood.
In \texttt{TMB}, the user writes C++ code to evaluate this negative log-likelihood function $\ell$.
A standard maximum likelihood approach is to optimise
\begin{equation}
L_\ell(\btheta) \triangleq \int_{\mathbb{R}^n} p(\y \, | \, \x, \btheta) \text{d}\x = \int_{\mathbb{R}^n} \exp(-\ell(\x, \btheta)) \text{d}\x \label{eq:flikelihood}
\end{equation}
with respect to $\btheta$ to find the maximum likelihood estimator (MLE) $\hat \btheta$.
Taking a superficially more Bayesian approach than above, instead of $\ell$, the user may instead write a function to evaluate the negative joint penalised log-likelihood given by
\begin{equation}
f(\x, \btheta) 
\triangleq - \log p(\y \, | \, \x, \btheta) p(\x, \btheta)
= \ell(\x, \btheta) - \log p(\x, \btheta),
\end{equation}
equivalent up to an additive constant to the negative log-posterior.
<!-- \begin{equation} -->
<!-- f(\x, \btheta)  -->
<!-- = - \log p(\y, \x, \btheta)  -->
<!-- = - \log p(\x, \btheta \, | \, \y) - C, -->
<!-- \end{equation} -->
<!-- where $C = \log p(\y)$ is the log evidence. -->
Using $f$ in place of $\ell$, then the penalised likelihood is proportional to the posterior marginal of $\btheta$
\begin{equation}
L_f(\btheta) 
\triangleq \int_{\mathbb{R}^n} \exp(-f(\x, \btheta)) \text{d}\x 
\propto \int_{\mathbb{R}^n} p(\x, \btheta \, | \, \y) \text{d}\x = p(\btheta \, | \, \y). \label{eq:1}
\end{equation}
Integrating out the random effects directly, as in Equation~\ref{eq:1} above, is usually intractable because $\x$ is high-dimensional, so \citet[Equation 3]{kristensen2016tmb} use a Laplace approximation $L^\star_f(\btheta)$ based instead upon integrating out a Gaussian approximation to the random effects.
This Laplace approximation is analogous to the INLA approximation $\tilde p(\btheta \, | \, \y)$ given in Section~\ref{sec:approxi}.
\begin{equation*}
f''_{\x\x}(\hat{\bm{\mu}}(\btheta), \btheta) = - \frac{\partial^2}{\partial \x^2} \log p(\y, \x, \btheta) \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)}
= - \frac{\partial^2}{\partial \x^2} \log p(\x \, | \, \btheta, \y) \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)}
= \hat{\bm{Q}}(\btheta).
\end{equation*}
Inference proceeds by optimising $L^\star_f(\btheta)$ via minimisation of
\begin{equation}
-\log L^\star_f(\btheta) \propto \frac{1}{2} \log \det (\hat{\bm{Q}}(\btheta)) + f(\hat{\bm{\mu}}(\btheta), \btheta)  \label{eq:nllaplace},
\end{equation}
where $\propto$ is used to mean proportional up to an additive constant.
The parameters of the Gaussian approximation (Equation~\ref{eq:gaussianx}), are found in terms of $f$ via $\hat{\bm{\mu}}(\btheta) = \arg \min_\x f(\x, \btheta)$ and $\hat{\bm{Q}}(\btheta) = f''_{\x\x}(\hat{\bm{\mu}}(\btheta), \btheta)$ and must be recomputed for each value of $\btheta$.
Obtaining $\hat{\bm{\mu}}(\btheta)$ is known as the inner optimisation step.

# Examples \label{sec:examples}

## The Naomi small-area estimation model

@eaton2019joint specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
Modelling data from multiple sources concurrently increases statistical power, and may mitigate the biases of any single source giving a more complete picture of the situation as well as prompting investigation into any conflicts.

**Prevalence component**
  Consider a country partitioned into areas $i = 1, \ldots, n$, with a simple random household survey of $m_i$ people is conducted in each area with $y_i$ HIV positive cases observed.
Cases may be modelled using a binomial logistic regression model
\begin{align}
y_i &\sim \text{Bin}(m_i, \rho_i), \\
\text{logit}(\rho_i) &\sim \mathcal{N}(\beta_\rho, \sigma_\rho^2)
\end{align}
where HIV prevalence $\rho_i$ is modelled by a Gaussian with mean $\beta_\rho$ and standard deviation $\sigma_\rho$.

**ANC component**
  Routinely collected data from pregnant women attending antenatal care clinics (ANCs) is another important source of information about the HIV epidemic.
If of $m^\text{ANC}_i$ women, $y^\text{ANC}_i$ are HIV positive, then an analogous binomial logistic regression model
\begin{align}
y^\text{ANC}_i &\sim \text{Bin}(m^\text{ANC}_i, \rho^\text{ANC}_i), \\
\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + b_i, \\
b_i &\sim \mathcal{N}(\beta_b, \sigma_b^2),
\end{align}
may be used to describe HIV prevalence amongst the sub-population of women attending ANCs.
Reflecting the fact that prevalence in ANCs is related but importantly different to prevalence in the general population, bias terms $b_i$ are used to offset ANC prevalence from HIV prevalence.

**ART component**
  The number of people receiving treatment at district health facilities $A_i$ also provides additional information about HIV prevalence, whereby districts with high prevalence are likely to have a greater number of people receiving treatment.
ART coverage, defined to be the proportion of PLHIV currently on ART on district $i$, is given by $\alpha_i = A_i / \rho_i N_i$, where $N_i$ is the total population of district $i$ and assumed to be fixed.
As such, ART coverage may also be modelled using a binomial logistic regression model
\begin{align}
A_i &\sim \text{Bin}(N_i, \rho_i \alpha_i), \\
\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2).
\end{align}

@@

**Appendix A**: Statistical modelling \newline
**Appendix B**: Supplementary tables and figures

>>>>

---
title: Simplified integrated nested Laplace approximation for extended latent Gaussian models
author:
  - name: Adam 
    surname: Howes
    email: ath19@ic.ac.uk
    label: e1
    addressLabel: A
    sepNext: "," # use "," or \and; the last author not need this field
  - name: Alex
    surname: Stringer
    email: alex.stringer@uwaterloo.ca
    label: e2
    addressLabel: B
affiliation:
  - label: A
    name: Department of Mathematics, Imperial College London
    authorsLabels: e1
  - label: B
    name: Department of Statistics and Actuarial Science, University of Waterloo
    authorsLabels: e2
abstract: |
  | Naomi is a spatial evidence synthesis model used by countries in sub-Saharan Africa to produce HIV epidemic indicators. Performing inference for Naomi is challenging because it is an extended latent Gaussian model (ELGM) and not amenable to inference via INLA as described by Rue et al (2009). We combine the simplified INLA approach of Wood (2020) with adaptive Gaussian hermite quadrature to enable inference for ELGMs. Using data from Malawi, we compare our inference method to other approaches. We provide an easy to use implementation as a part of the `aghq` R package, allowing more flexible use of the INLA method for a broader class of models, including any with a TMB template.
keyword-subclass: | 
 \begin{keyword}[class=MSC2020] % It must be define for aap, aop, aos journals. For aoas, sts is not used
 \kwd[Primary ]{00X00}
 \kwd{00X00}
 \kwd[; secondary ]{00X00}
 \end{keyword}
keywords: 
  - Spatial statistics
  - INLA

predefined-theoremstyle: true # use in section Environments for Axiom, Theorem, etc
bibliography: citations.bib
biblio-style: imsart-nameyear # alternative: imsart-number
output:
  rticles::ims_article:
    journal: aoas # aap, aoas, aop, aos, sts. See documentation
    toc: false # Please use for articles with 50 pages and more
    includes:
      in_header: preamble.tex
---

@@

We are motivated by a challenging inference problem in HIV surveillance.
Accurate estimates of HIV indicators are required to mount effective public health response to the epidemic.
The Naomi small-area estimation model [@eaton2021naomi] synthesises data from multiple sources to estimate HIV prevalence, HIV incidence, and coverage of antiretroviral treatment (ART) at a district-level.
Software has been developed for the model allowing countries to use their data to generate estimates in a yearly process supported UNAIDS.
As such, the inference method used must be suitable to run in production, ruling out prohibitively slow Markov chain Monte Carlo approaches.
Furthermore, as Naomi falls into the class of extended latent Gaussian models (ELGMs) [@stringer2022fast] it cannot be fit using the integrated nested Laplace approximation (INLA)  method [@rue2009approximate] as implemented by the `R-INLA` \textsf{R} package.
Instead, inference is conducted using the Template Model Builder (`TMB`) \textsf{R} package [@kristensen2016tmb], which implements a variety of empirical Bayes inference, and is gaining popularity in spatial statistics as a flexible alternative to `R-INLA` [@osgoodzimmerman2021statistical].

@@

INLA is an approximate Bayesian inference method based on the Laplace approximation and numerical integration.
INLA is designed for use with latent Gaussian models (LGMs) of the form
\begin{alignat}{2}
&\text{(Observations)}     &        y_i &\sim p(y_i \, | \, x_i, \btheta), \quad i = 1, \ldots, n, \label{eq:data} \\
&\text{(Latent field)}     &        \x &\sim \mathcal{N}(\x \, | \, \mathbf{0}, \mathbf{Q}(\btheta)^{-1}), \label{eq:process} \\
&\text{(Parameters)}       & \qquad \btheta &\sim p(\btheta), \label{eq:parameters}
\end{alignat}
where $\text{dim}(\y) = \text{dim}(\x) = n$ and $\text{dim}(\btheta) = m$, and $m < n$.
The joint posterior of $(\x, \btheta)$ is given by
\begin{equation}
  p(\x, \btheta \, | \, \y) 
  \propto p(\btheta) |\mathbf{Q}(\btheta)|^{n/2} \exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right). \label{eq:lgmjointposterior}
\end{equation}
Rather than approximating the above full posterior, the INLA method instead approximates the posterior marginals of each latent random variable $x_i$ and parameter $\theta_j$ given by
\begin{align}
  p(x_i \, | \, \y) &= \int p(x_i, \btheta \, | \, \y) \text{d} \btheta = \int p(x_i \, | \, \btheta, \y) p(\btheta \, | \, \y) \text{d}\btheta, \quad i = 1, \dots, n, \label{eq:inla1} \\
  p(\theta_j \, | \, \y) &= \int p(\btheta \, | \, \y) \text{d}\btheta_{-j} \quad j = 1, \ldots, m. \label{eq:inla2}
\end{align}
An approximation is made to each of the two quantities, $p(\btheta \, | \, \y)$ and $p(x_i \, | \, \btheta, \y)$, nested inside the above integrals: (i) $p(\btheta \, | \, \y) \approx \tilde p(\btheta \, | \, \y)$ and (ii) $p(x_i \, | \, \btheta, \y) \approx \tilde p(x_i \, | \, \btheta, \y)$, which we discuss in turn below.

@@

The posterior marginal of the parameters $p(\btheta \, | \, \y)$ appears in both Equations \eqref{eq:inla1} and \eqref{eq:inla2}.
This distribution is approximated by $\tilde p(\btheta \, | \, \y)$ and represented by a set of $K$ integration points $\{ \btheta^{(k)} \}$ and area-weights $\{ \Delta^{(k)} \}$.
The first step is to rewrite $p(\btheta \, | \, \y)$ as
\begin{equation}
  p(\btheta \, | \, \y) 
  = \frac{p(\x, \btheta \, | \, \y)}{p(\x \, | \, \btheta, \y)}
  \propto \frac{p(\y, \x, \btheta)}{p(\x \, | \, \btheta, \y)}. \label{eq:hyperpost}
\end{equation}
Approximation (i) then uses a Gaussian approximation to the denominator of Equation \ref{eq:hyperpost} given by
\begin{equation}
p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y) \triangleq \mathcal{N}(\x \, | \, \hat{\bm{\mu}}(\btheta), \hat{\mathbf{Q}}(\btheta)^{-1}). \label{eq:gaussianx}
\end{equation}
This approximation is accurate as the Gaussian prior on the latent field $\x$ makes the posterior distribution, given by
\begin{equation}
p(\x \, | \, \btheta, \y) \propto
\exp \left( - \frac{1}{2} \x^\top \mathbf{Q}(\btheta) \x + \sum_{i = 1}^n \log p(y_i \, | \, x_i, \btheta) \right),
\end{equation}
close to being Gaussian because $\y$ is generally not that informative and the observation distribution $p(\y \, | \, \x, \btheta)$ is usually well-behaved [@blangiardo2015spatial].
<!-- If $\y$ is uninformative about $\x$ then how can we hope for useful inference? Perhaps what the author (Marta) meant is that the Gaussian distribution is sufficiently flexible to accommodate the updates made by $\y$. -->
As $p(\btheta \, | \, \y)$ does not depend on $\x$, any value may be chosen to evaluate the right hand side of Equation \ref{eq:hyperpost}.
As such, taking $\x = \hat{\bm{\mu}}(\btheta)$, the value where the Gaussian approximation is most accurate, gives the final approximation as
\begin{equation}
\tilde p(\btheta \, | \, \y) \propto \frac{p(\y \, | \, \x, \btheta) p(\x \, | \, \btheta) p(\btheta)}{p_G(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)} = \frac{p(\y, \hat{\bm{\mu}}(\btheta), \btheta)}{\det(\hat{\bm{Q}}(\btheta))^{1/2}}, \label{eq:hypermarginal}
\end{equation}
where the final equality is because $p_G(\x \, | \, \btheta, \y)$ is evaluated at its mode $\hat{\bm{\mu}}(\btheta)$.

@@

Having used used the Gaussian approximation $p(\x \, | \, \btheta, \y) \approx p_G(\x \, | \, \btheta, \y)$ in Section \ref{sec:approxi} above, a natural approach, and that taken by \citet{rue2007approximate}, is to marginalise this distribution directly to obtain
\begin{equation}
\tilde p(x_i \, | \, \btheta, \y) = \mathcal{N}(x_i \, | \, \hat{\mu}_i(\btheta), 1 / \hat{q}_i(\btheta)),
\end{equation}
where the marginal mean $\hat{\mu}_i(\btheta)$ and precision $\hat{q}_i(\btheta)$ are recovered directly from $\hat{\bm{\mu}}(\btheta)$ and $\hat{\bm{Q}}(\btheta))$ respectively.
Although this approximation is fast, it tends not to be accurate, as it involves evaluating the Gaussian approximation away from its mode.
As a result, although this method is available in \texttt{R-INLA} it is generally not advised.
Instead, \citet{rue2009approximate} propose two methods, a Laplace approximation and a simplified version which is less computationally demanding.
The full Laplace approximation is
\begin{align}
  p(x_i \, | \, \btheta, \y) &= p(\x \, | \, \btheta, \y) \times \frac{1}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\
  &= \frac{p(\x, \btheta \, | \, \y)}{p(\btheta \, | \, \y)} \times \frac{1}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\
  &\propto \frac{p(\x, \btheta \, | \, \y)}{p(\x_{-i} \, | \, x_i, \btheta, \y)} \\
  &\approx \frac{p(\x, \btheta \, | \, \y)}{p_G(\x_{-i} \, | \, x_i, \btheta, \y)} \Big\rvert_{\x_{-i} = \hat{\bm{\mu}}_{-i}(x_i, \btheta)} 
  = \tilde p_{LA}(x_i \, | \, \btheta, \y), \label{eq:laplaceapproximation}
\end{align}
where $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ is the Gaussian approximation to $\x_{-i} \, | \, x_i, \btheta, \y$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is its modal configuration.\footnote{Note that $\hat{\bm{\mu}}(\btheta)$ is the mode of the Gaussian approximation to the full latent field given $\btheta$ and $\hat{\bm{\mu}}_{-i}(x_i, \btheta)$ is not the same as $\hat{\bm{\mu}}_{-i}(\btheta)$.}
The set of distributions $\{p(\x_{-i} \, | \, x_i, \btheta, \y)\}_{i = 1}^n$ are usually reasonably Gaussian so this approximation tends to work well.
However, the Gaussian approximation $p_G(\x_{-i} \, | \, x_i, \btheta, \y)$ must be recomputed for each value of $x_i$ and $\btheta$, which is often computationally prohibitive.
Therefore, two modifications to Equation \eqref{eq:laplaceapproximation} are proposed by \citet{rue2009approximate} to reduce the computational cost:
\begin{enumerate}
\item
Avoiding having to find the mode via optimisation by using the approximation $\hat{\bm{\mu}}_{-i}(x_i, \btheta) \approx \mathbb{E}_{p_G(\x \, | \, \btheta, \y)}(\x_{-i} \, | \, x_i)$
\item
As only those $x_j$ close to $x_i$ should have an impact on the marginal of $x_i$, then by selecting some subset $R_i(\btheta)$ of nodes $j$ to impact $j$ the matrix which needs to be factorised can be reduced in dimension to be $| R_i(\btheta) | \times | R_i(\btheta) |$ rather than $n \times n$
\end{enumerate}

@@

## Template Model Builder \label{sec:tmb}

`TMB` [@kristensen2016tmb] is an \textsf{R} package for fitting random effect models, also known as latent variable models, hierarchical models or a host of other names.
In `TMB` inference is based upon optimisation of a target function.
This makes it very flexible, and able to handle non-linear, non-Gaussian random effect models.

The approach of `TMB` is inspired by the AD Model Builder (ADMB) package [@fournier2012ad].
The "AD" in ADMB is automatic differentiation, a technique for calculating derivates of functions by repeated application of the chain rule.
AD is popular in machine learning [@baydin2017automatic], for example as the basis for backpropagation algorithm and is beginning to gain popularity in statistics, including as a part of Stan [@carpenter2017stan].
`TMB` uses the derivatives from AD for multiple purposes including calculation of the Hessian used in Gaussian approximations and for numerical optimisation routines.

Consider unobserved latent random effects $\x \in \mathbb{R}^n$ and parameters $\btheta \in \mathbb{R}^m$.\footnote{\citet{kristensen2016tmb] use the notation $u$ for random effects and $\theta$ for parameters. We aim for consistency with Section \ref{sec:inla}.}
Let $\ell(\x, \btheta) \triangleq - \log p(\y \, | \, \x, \btheta)$ be the negative joint log-likelihood.
In `TMB`, the user writes C++ code to evaluate this negative log-likelihood function $\ell$.
A standard maximum likelihood approach is to optimise
\begin{equation}
L_\ell(\btheta) \triangleq \int_{\mathbb{R}^n} p(\y \, | \, \x, \btheta) \text{d}\x = \int_{\mathbb{R}^n} \exp(-\ell(\x, \btheta)) \text{d}\x \label{eq:flikelihood}
\end{equation}
with respect to $\btheta$ to find the maximum likelihood estimator (MLE) $\hat \btheta$.
Taking a superficially more Bayesian approach than above, instead of $\ell$, the user may instead write a function to evaluate the negative joint penalised log-likelihood given by
\begin{equation}
f(\x, \btheta) 
\triangleq - \log p(\y \, | \, \x, \btheta) p(\x, \btheta)
= \ell(\x, \btheta) - \log p(\x, \btheta),
\end{equation}
equivalent up to an additive constant to the negative log-posterior.
<!-- \begin{equation} -->
<!-- f(\x, \btheta)  -->
<!-- = - \log p(\y, \x, \btheta)  -->
<!-- = - \log p(\x, \btheta \, | \, \y) - C, -->
<!-- \end{equation} -->
<!-- where $C = \log p(\y)$ is the log evidence. -->
Using $f$ in place of $\ell$, then the penalised likelihood is proportional to the posterior marginal of $\btheta$
\begin{equation}
L_f(\btheta) 
\triangleq \int_{\mathbb{R}^n} \exp(-f(\x, \btheta)) \text{d}\x 
\propto \int_{\mathbb{R}^n} p(\x, \btheta \, | \, \y) \text{d}\x = p(\btheta \, | \, \y). \label{eq:1}
\end{equation}
Integrating out the random effects directly, as in Equation \ref{eq:1} above, is usually intractable because $\x$ is high-dimensional, so \citet[Equation 3]{kristensen2016tmb} use a Laplace approximation $L^\star_f(\btheta)$ based instead upon integrating out a Gaussian approximation to the random effects.
This Laplace approximation is analogous to the INLA approximation $\tilde p(\btheta \, | \, \y)$ given in Section \ref{sec:approxi}.
\begin{equation*}
f''_{\x\x}(\hat{\bm{\mu}}(\btheta), \btheta) = - \frac{\partial^2}{\partial \x^2} \log p(\y, \x, \btheta) \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)}
= - \frac{\partial^2}{\partial \x^2} \log p(\x \, | \, \btheta, \y) \Big\rvert_{\x = \hat{\bm{\mu}}(\btheta)}
= \hat{\bm{Q}}(\btheta).
\end{equation*}
Inference proceeds by optimising $L^\star_f(\btheta)$ via minimisation of
\begin{equation}
-\log L^\star_f(\btheta) \propto \frac{1}{2} \log \det (\hat{\bm{Q}}(\btheta)) + f(\hat{\bm{\mu}}(\btheta), \btheta)  \label{eq:nllaplace},
\end{equation}
where $\propto$ is used to mean proportional up to an additive constant.
The parameters of the Gaussian approximation (Equation \ref{eq:gaussianx}), are found in terms of $f$ via $\hat{\bm{\mu}}(\btheta) = \arg \min_\x f(\x, \btheta)$ and $\hat{\bm{Q}}(\btheta) = f''_{\x\x}(\hat{\bm{\mu}}(\btheta), \btheta)$ and must be recomputed for each value of $\btheta$.
Obtaining $\hat{\bm{\mu}}(\btheta)$ is known as the inner optimisation step.

# The Naomi small-area estimation model

@eaton2019joint specify a joint model linking small-area estimation models of HIV prevalence from household surveys, HIV prevalence from antenatal care clinics, and antiretroviral therapy (ART) coverage from routine health data collection.
This model forms the basis of the Naomi small-area estimation model [@eaton2021naomi]. 
Modelling data from multiple sources concurrently increases statistical power, and may mitigate the biases of any single source giving a more complete picture of the situation, as well as prompting investigation into any data conflicts.
The model is described by three components, as follows.

### Prevalence component

Consider a country partitioned into areas $i = 1, \ldots, n$.
A simple random household survey of $m_i$ people is conducted in each area, and $y_i$ HIV positive cases are observed.
Cases may be modelled using a binomial logistic regression model
\begin{align}
y_i &\sim \text{Bin}(m_i, \rho_i), \\
\text{logit}(\rho_i) &\sim \mathcal{N}(\beta_\rho, \sigma_\rho^2)
\end{align}
where HIV prevalence $\rho_i$ is modelled by a Gaussian with mean $\beta_\rho$ and standard deviation $\sigma_\rho$.

### ANC component

Routinely collected data from pregnant women attending antenatal care clinics (ANCs) is another important source of information about the HIV epidemic.
If, of $m^\text{ANC}_i$ women, $y^\text{ANC}_i$ are HIV positive, then an analogous binomial logistic regression model
\begin{align}
y^\text{ANC}_i &\sim \text{Bin}(m^\text{ANC}_i, \rho^\text{ANC}_i), \\
\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + b_i, \\
b_i &\sim \mathcal{N}(\beta_b, \sigma_b^2),
\end{align}
may be used to describe HIV prevalence amongst the sub-population of women attending ANCs.
Reflecting the fact that prevalence in ANCs is related but importantly different to prevalence in the general population, bias terms $b_i$ are used to offset ANC prevalence from HIV prevalence.

### ART component

The number of people receiving treatment at district health facilities $A_i$ also provides additional information about HIV prevalence.
Districts with high prevalence are likely to have a greater number of people receiving treatment, and vice versa.
ART coverage, defined to be the proportion of PLHIV currently on ART on district $i$, is given by $\alpha_i = A_i / \rho_i N_i$, where $N_i$ is the total population of district $i$ and assumed to be fixed.
As such, ART coverage may also be modelled using a binomial logistic regression model
\begin{align}
A_i &\sim \text{Bin}(N_i, \rho_i \alpha_i), \\
\text{logit}(\alpha_i) &\sim \mathcal{N}(\beta_\alpha, \sigma_\alpha^2).
\end{align}

# Results \label{sec:results}

# Conclusions \label{sec:conclusions}

@@

**Appendix A**: Supporting information
